---
output: md_document
---
# Reference classes OOP

Reference classes are a Java-like class system (no generics!) implementation in R. 

Reference classes use pass-by-reference logic that is different from the usual pass-by-value logic R uses. This means when we assign `a=b` both `a` and `b` point to the same object, and changing `a` will also modify `b` and vica versa. 

## Creating classes

Similar syntax as S4, but now slots are called fields, and the `setRefClass` returns a constructor object. 

```{r}
Seq <- setRefClass("Seq",
                   fields = list(
                     name = "character",
                     alphabet = "character",
                     sequence = "character"))
```

We then use the constuctor to create objects:

```{r}
a = Seq(name="My seq", alphabet=c("A", "T"), sequence="ATTTTAAAAAA")
a
```

## Defining methods

Methods are defined either with the class, or later by modifying the constructor object. 

```{r}
Seq <- setRefClass("Seq",
                   fields = list(
                     name = "character",
                     alphabet = "character",
                     sequence = "character"),
                   methods = list(
                     comp = function() {
                       "Complements the (DNA) sequence" ## inline docs
                       sequence <<- chartr("ACGT","TGCA",.self$sequence)
                       name <<- paste(.self$name, "-- complemented")                     
                     }
                     ## there can be more, of course
                     ))
```
 
A couple of things to note here:

- accessing fields and calling methods is done with the `$` operator.
- the current object can be referred to in a method by the reserved name `.self`.
- Changing fields of an object within methods needs to be done with the `<<-` operator.

Calling a method on the object will modify it:

```{r}
s <- Seq$new(name="foo", sequence="GATCATCA")
s
s$sequence
s$comp()
s$sequence
```

Objects are copied by reference and not value:

```{r}
s2 = s
s$sequence
s2$sequence
s$sequence = "ATTTA"
s$sequence
s2$sequence
```

# Summary

Reference classes are suitable for objects that are dynamically tracked by all the code: GUI components, read-only access to files (streams, data bases), internet resources, editing facilities, ...

Because of their pass-by-reference semantics they are not appropriate as a replacement for S3 and S4 as this would lead to a lot of unanticipated side effects. 

