---
output: md_document
---
# S3 OOP

Many of the commonly used R functions use S3 in the background, such as `plot`, `summary`, `mean`, ... 

Lets see how `summary` works. Note that it produces different outputs for different _types_ of inputs:

```{r}
df = data.frame(a=1:5, b=6:10)
vect = 1:5

class(df)
class(vect)

summary(df)
summary(vect)
```

So how does the code for `summary` look like, does it have a lot of if-else statements to provide different outputs for different types of inputs? No, it uses S3 OOP to achieve that. 

```{r}
summary

methods(summary)
```

The algorithm by which an appropriate method is called is called *dispatching*:

1. The line of code is `UseMethod("summary")` is the dispatch function call, and we call `summary` a *generic function*
2. The call to this function finds out the class of the first argument (ie `class(object)`), and based on it calls the appropriate implementation:
    1. If a function with name `summary.<class name>` exists, it will call that one
    2. If such a function doesn't exist, it calls `summary.default`

Therefore we are effectively calling:
```{r}
summary.data.frame(df)  # same as summary(df) because class(df) == "data.frame"
summary.default(vect)  # same as summary(vect) because class(vect) == "integer"
```

What if we wanted `summary(vect)` to output something different? We just define a new method. 

```{r}
summary.integer = function(object, ...){
  message("This is our own version of summary for integers!")
}
methods("summary")
summary(vect)
```

Therefore, generics and dispatching enable us to re-implement some of the well-known R functions for any data type, including our own - which we'll cover next. 

### Introducing our own classes

We can also create our own data types (classes). Lets says we want to wrap two pieces of information:

- DNA sequence name
- DNA sequence  

into a single new data type. And, we want to implement our own version of ```summary``` for this new data type. 

To create a new object of our class `MyDNASeq` we just modify the `class` attribute of any base R object, e.g. a list. 

```{r}
dna = list(name="Our sequence name", sequence="ATGGATGACGATG")
class(dna) = "MyDNASeq"
dna
```

At this point we created a new object `dna` with class `MyDNASeq`. We can now go ahead and plug in our implementation for `summary`:

```{r}
summary(dna)

summary.MyDNASeq = function(object, ...){
  message("This is a DNA sequence of length ", nchar(dna$sequence))
}

summary(dna)
```

You can also introduce generics by creating a function that has `UseMethod("<name of the generic>")` as the only code. 

### Generics from primitives

All primitve functions are implicily generics, e.g. `length`:

```{r}
length
methods(length)
```

So we can define methods as usual:

```{r}

length(dna) # before defining our implementation

length.MyDNASeq = function(x){
  nchar(x$sequence)
}

length(dna)
```

Here we redefined `length()` to return the number of characters in our DNA sequence.  

### Exercise: create a plotting function for our "MyDNASeq" type

Calling `plot()` on our data doesn't work. Create your own implementation of `plot()` for `MyDNASeq`. You can put any meaningful code, e.g. plot a histogram of the distribution of DNA letters.  

```{r, error=TRUE}
plot(dna)
```

<!--
### Solution

```{r}
plot.MyDNASeq = function(x, y, ...){
  letters = table(strsplit(x$sequence,""))
  barplot(letters)
}

plot(dna)
```
-->
## Issues with S3

S3 is very informal, with very little consistency checking, so things can easily go wrong:

```{r, error=TRUE}
class(dna) = "lm" # dna is now a linear model?!
```

R lets us set any `class` atribute for any object. However, the object is invalid and we'll get and error if we try to use it:

```{r, error=TRUE}
summary(dna)
```

## S3 summary

- New types are created by setting the `class` property of any existing base R object, usually a `list`.
- New methods are created by naming convention `generic.className`.
- Easy to make errors, e.g. mistype class name, assing nonsensical class names, or unknowingly override a method if you use `.` in your function names. 
- No consisency checking - no formal guarantee that two objects of same class will have the same properties

