---
title: "An Introduction to Machine Learning with R"
output:
  html_document:
    toc: yes
    toc_float: true
    self_contained: true	
---

```{r setup, echo=FALSE}
source("./src/setup.R")
```

# An Introduction to Machine Learning with R

This introductory workshop on machine learning with R is aimed at
participants who are not experts in machine learning (introductory
material will be presented as part of the course), but have some
familiarity with scripting in general and R in particular. The
workshop will offer a hands-on overview of typical machine learning
applications in R, including unsupervised (clustering, such as
hierarchical and k-means clustering, and dimensionality reduction,
such as principal component analysis) and supervised (classification
and regression, such as K-nearest neighbour and linear regression)
methods. We will also address questions such as model selection using
cross-validation. The material will have an important hands-on
component and participants will need to bring a computer with R 3.4.1
pre-installed. All the course material will be made available online
and package instructions will be circulated in advance.


# Objectives and pre-requisites

- The course aims at providing an accessible introduction to various
  machine learning methods and applications in R. The core of the
  courses focuses on unsupervised and supervised methods. 

- The course contains numerous expercices to provide amble opportunity
  to apply the newly learnt material.
  
- Participants are expected to be familiar with the R syntax and basic
  plotting functionality.

- At the end of the course, the participants are anticipated to be
  able to apply the methods learnt, as well as feel confident enough
  to explore and apply new methods.

# Why R?

R is one of the major languages for data science currently
available. It provides excellent visualisation features, which is
essential to explore the data before submitting it to any automated
learning, as well as assessing the results of the learning
algorithm. Many R package
for [machine learning](https://cran.r-project.org/) are availble of
the shelf and many modern methods in statical learning are implemented
in R as part of their development.

Nowadays, there are however other viable alternatives that benefit
from similar advantages. If we consider python, for example,
the [scikit-learn](http://scikit-learn.org/stable/index.html) software
provides all tools that we will present in this course.

# Overview of machine learning (ML)


In **supervised learning** (SML), the learning algorithm is presented
with labelled example inputs, where the labels indicate the desired
output. SML itself is composed of **classification**, where the output
is categorical, and **regression**, where the output is numerical.

In **unsupervised learning** (UML), no labels are provided, and the
learning algorithm focuses solelt on detecting structure in unlabelled
input data.

Note that there are also **semi-supervised learning** approaches that
use labelled data to inform unsupervised learning on the unlabelled
data to identify and annotate new classes in the dataset (also called
novely detection). 

**Reinforcement learning**, the learning algorithm performs a task
using feedback from operating in a real of synthetic environment.


## Example data

- *Observations*, *examples* or simply *data points* along the rows
- *Features* or *variables* along the columns

Using the *iris* data as an example, for UML, we would have 4 features
for each unlabelled example.

```{r irisuml, echo=FALSE}
data(iris)
knitr::kable(head(iris[, 1:4]))
```

The same dataset used in the context of SML contains an additional
column of labels, documenting the outcome or class of each example.

```{r irissml, echo=FALSE}
knitr::kable(head(iris[, c(5, 1:4)]))
```

The different datasets that are used throughout the course are
collected and briefly described in the short *Data* chapter.

# Content

- Introduction (this chapter)
- Unsupervised machine learning
- Supervised machine learing
- Conclusing remarks and references
