---
title: "An Introduction to Machine Learning with R: Unsupervised Learning"
output:
  html_document:
    toc: yes
    toc_float: true
    self_contained: true	
---

```{r setup, echo=FALSE}
source("./src/setup.R")
```

# Unsupervised learning

In **unsupervised learning** (UML), no labels are provided, and the
learning algorithm focuses solelt on detecting structure in unlabelled
input data. One generally differentiates between

- **Clustering**, where the goal is to find homogeneous subgroups
  within the data; the grouping is based on distance between
  observations.
  
- **Dimensional reduction**, where the goal is to identify patterns in
  the features of the data. Dimensionality reduction is often used to
  facilitate visualisation of the data, as well as a pre-processing
  method before supervised learning.

UML presents specific challenges and benefits:
- there is no single goal in UML
- there is generally much more unlabelled data available than labelled
  data.

# k-means clustering

In R: 

```{r}
stats::kmeans(x, centers = 3, nstart = 20)
```

# Hierarchical clustering

# Pre-processing

# Principal component analysis (PCA)

# t-SNE



