---
title: "Searching GEO"
author: "Brian High"
date: "2/3/2015"
output:
  ioslides_presentation:
    fig_caption: yes
    fig_retina: 1
    keep_md: yes
    smaller: yes
---

## Setting up some options

Let's first turn on the cache for increased performance and improved styling.

```{r, cache=FALSE}
# Set some global knitr options
library("knitr")
opts_chunk$set(tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=60), 
               cache=FALSE, messages=FALSE)
```

## Prepare for HW2

We will need to query the GEOmetadb database. Today we will explore this 
database and practice various ways to query it.

## Load the `GEOmetadb` package

First we load the `GEOmetadb` library.

```{r}
suppressMessages(library(GEOmetadb))
```

Let's also view the available methods.

```{r}
ls("package:GEOmetadb")
```

## Download the GEO database

We should have already downloaded this database when viewing the lecture slides.

```{r}
## This will download the entire database, so can be slow
if(!file.exists("GEOmetadb.sqlite"))
{
  # Download database only if it's not done already
  getSQLiteFile()
}
```

## List tables with `SQL`

In `SQL`, you can query the database structure with ordinary `SQL` commands.

```{r}
geo_con <- dbConnect(SQLite(),'GEOmetadb.sqlite')
dbGetQuery(geo_con, "SELECT name FROM sqlite_master WHERE type='table';")
```

## List `gse` fields with `SQL`

The `PRAGMA` command is a standard `SQLite` command.

```{r}
dbGetQuery(geo_con, "PRAGMA table_info(gse);")
```

## List tables with `dbListTables`

Instead of using `SQL` commands, we can list tables and fields with functions
from the `GEOmetadb` package.

```{r}
geo_con <- dbConnect(SQLite(),'GEOmetadb.sqlite')
dbListTables(geo_con)
```

```{r}
dbListFields(geo_con, 'gse')
```

## Explore `gse`

```{r}
columnDescriptions()[1:5,]
```

## Load library `data.table`

This will provide us with some practice querying with data.table.

```{r}
suppressMessages(library(data.table))
```

## Explore `gse` with `data.table`

```{r}
cd <- as.data.table(columnDescriptions())
cd[TableName=="gse", FieldName]
```

## Load library `pander`

This will give us the `pandoc.table` function which will make nicer table 
listings.

```{r}
suppressMessages(library(pander))
```

## List `gse` columns with `pandoc.table`

```{r}
gsefields <- as.data.frame(
    cd[TableName=="gse" & 
           FieldName %in% c("gse","title","pubmed_id","summary","contact")])
pandoc.table(gsefields, style = "grid")
```

## Explore `gpl`

```{r}
cd[TableName=="gpl", FieldName]
```

## Explore columns in `gpl`

```{r}
gplfields <- as.data.frame(
    cd[TableName=="gpl" & 
           FieldName %in% c("gpl","organism","manufacturer")])
pandoc.table(gplfields, style = "grid")
```

## Explore `gse_gpl`

```{r}
cd[TableName=="gse_gpl", FieldName]
```

## Explore columns in `gse_gpl`

```{r}
gse_gplfields <- as.data.frame(cd[TableName=="gse_gpl"])
pandoc.table(gse_gplfields, style = "grid")
```

## List some data

Let's look at some records in `gse`. What does a "contact" look like?

```{r}
query <- "SELECT contact FROM gse LIMIT 1;"
res <- dbGetQuery(geo_con, query)
strsplit(res$contact, '\t')
```

## Find manufacturer data

Query the manufacturers with a `SQL` command, listed with `data.table`...

```{r}
manu <- data.table(dbGetQuery(geo_con, 
    "SELECT DISTINCT manufacturer FROM gpl ORDER BY manufacturer ASC;"))
manu[,list(length(manufacturer)), by=manufacturer]
```

## Our `SQL` command

We just wanted a list of manufacturers so the `SQL` query is:

```
SELECT DISTINCT manufacturer FROM gpl 
ORDER BY manufacturer ASC;
```

However, since we also grouped `by=manufacturer` in our `data.table`, we could 
have simply used the `SQL` query:

```
SELECT DISTINCT manufacturer FROM gpl;
```
Let's try that...

## Find manufacturer data

Query the manufacturers with a simpler `SQL` command ... grouping with `by` and 
ordering with `setkey` in `data.table`...

```{r}
manu <- data.table(dbGetQuery(geo_con, 
            "SELECT DISTINCT manufacturer FROM gpl;"))
setkey(manu, manufacturer)
manu[,list(length(manufacturer)), by=manufacturer]
```


## Finding data with a `JOIN`

To get supplementary file names ending with `CEL.gz` from only manufacturer 
Affymetrix...

```
SELECT 
        gpl.bioc_package, 
        gsm.title, 
        gsm.series_id, 
        gsm.gpl, 
        gsm.supplementary_file 
    FROM gsm 
    JOIN gpl ON gsm.gpl=gpl.gpl 
    WHERE gpl.manufacturer='Affymetrix' 
        AND gsm.supplementary_file like '%CEL.gz';
```

## Now let's run that query

```{r}
res <- dbGetQuery(geo_con, 
    "SELECT 
        gpl.bioc_package, 
        gsm.title, 
        gsm.series_id, 
        gsm.gpl, 
        gsm.supplementary_file 
    FROM gsm JOIN gpl ON gsm.gpl=gpl.gpl 
    WHERE gpl.manufacturer='Affymetrix' 
        AND gsm.supplementary_file like '%CEL.gz';")
head(res)
```
