# Wrap-up

```{r, echo = FALSE, warning = FALSE}
suppressPackageStartupMessages(library("gplots"))
suppressPackageStartupMessages(library("pRoloc"))
suppressPackageStartupMessages(library("pRolocdata"))
suppressPackageStartupMessages(library("lattice"))
suppressPackageStartupMessages(library("msdata"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("mzID"))
suppressPackageStartupMessages(library("mzR"))
suppressPackageStartupMessages(library("RforProteomics"))
suppressPackageStartupMessages(library("gridExtra"))
suppressPackageStartupMessages(library("lattice"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("readxl"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("hexbin"))
suppressPackageStartupMessages(library("magrittr"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("knitr"))
suppressPackageStartupMessages(library("lattice"))
suppressPackageStartupMessages(library("RColorBrewer"))
suppressPackageStartupMessages(library("mzR"))
suppressPackageStartupMessages(library("gridExtra"))
suppressPackageStartupMessages(library("lattice"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("multtest"))
suppressPackageStartupMessages(library("limma"))
suppressPackageStartupMessages(library("msmsTests"))
suppressPackageStartupMessages(library("MSnID"))
suppressPackageStartupMessages(library("qvalue"))
suppressPackageStartupMessages(library("multtest"))
library("BiocStyle")
```

## References and resources

* [Visualisation of proteomics data using R and Bioconductor](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4510819/)
* [Using R and Bioconductor for proteomics data analysis](http://arxiv.org/pdf/1305.6559v1.pdf)
* `RforProteomics`: http://bioconductor.org/packages/RforProteomics
* [R/Bioconductor work-flow](http://bioconductor.org/help/workflows/proteomics/)

## Other relevant packages/pipelines

- Analysis of post translational modification with `r Biocpkg("isobar")`.
- Processing and analysis or isobaric tagging mass spectrometry with
  `r Biocpkg("isobar")` and `r Biocpkg("MSnbase")`.
- Analysis of spatial proteomics data with `r Biocpkg("pRoloc")`.
- Analysis of MALDI data with the `r Biocpkg("MALDIquant")` package.
- Access to the Proteomics Standard Initiative Common QUery InterfaCe
  with the `r Biocpkg("PSICQUIC")` package.
- `r Biocpkg("Cardinal")`: A mass spectrometry imaging toolbox for
  statistical analysis.
- `r CRANpkg("protViz") `: Visualising and Analysing Mass Spectrometry
  Related Data in Proteomics
- `r CRANpkg("aLFQ")`: Estimating Absolute Protein Quantities from
  Label-Free LC-MS/MS Proteomics Data.
- `r CRANpkg("protiq")`: Protein (identification and) quantification
  based on peptide evidence.
- `r Biocpkg("MSstats")`: Protein Significance Analysis in DDA, SRM
  and DIA for Label-free or Label-based Proteomics Experiments


### DIA

- Analysis of label-free data from a Synapt G2 (including ion
  mobility) with `r Biocpkg("synapter")`.
- `r Biocpkg("SWATH2stats")`: Transform and Filter SWATH Data for
  Statistical Packages and
- `r Biocpkg("specL")`: Prepare Peptide Spectrum Matches for Use in
  Targeted Proteomics
- `r Biocpkg("SwathXtend")`: SWATH extended library generation and
  statistical data analysis


### Session info

```{r}
sessionInfo()
```
