# A short microarray analysis

## Create an `marray` list

We are going to use `read.csv` to import our comma-separated spreadsheet into `R`.
Setting `row.names = 1` uses the first column of the imported spreadshhet 
to define the row names of the resulting data frame. 

```{r exp}
exp <- read.csv("MAdata1.csv", row.names = 1)
head(exp)
```

Because we want to store our expression data in a `matrix`, 
we explicitly coerce the `data.frame` into the adequate class.

```{r expmat}
class(exp)
exp <- as.matrix(exp)
class(exp)
```


We use the same call to create the sample and feature meta data, 
which are directly represented by the desired `data.frame` class.

```{r metadata}
smeta <- read.csv("smeta1.csv", row.names = 1)
fmeta <- read.csv("fmeta1.csv", row.names = 1)
class(smeta)
```

We now only need to create our final `marray` list as shown in the previous exercise.

```{r marraylist, tidy = FALSE}
marray <- list(expression = exp,
               featuremeta = fmeta,
               samplemeta = smeta)
str(marray)
```

To avoid repeating the above every time we use `R` to study this data, 
we `save` it on disk as a binary representation and verify that we can 
`load` it again.

```{r savemarray}
save(marray, file = "marray.rda")
rm(list = ls()) ## clean working environment
load("marray.rda")
str(marray)
```

The feature meta data has a `bh` column with FDR adjusted p-values using 
the Benjamini Hochberg method. Let's verify the number of significant 
values using 0.05 as threshold and count how many pass our threshold.


```{r marraynamesandde}
names(marray$featuremeta)
de <- marray$featuremeta[, "bh"] < 0.05
class(de)
table(de)
```
## Exploratory data analysis

A boxplot showing the distribution of expression data for each sample.

```{r marraybxplot}
boxplot(marray$expression)
```

A scatter plot comparing the expression values of the first and fourth samples

```{r expplot}
plot(marray$expression[, 1],
     marray$expression[, 4])
```

Let's tune the above scatter plot. To reduce typing, the expression data is stored in the `exp` variable.
We customise the plot with axes labels, a title, a grid, and a diagonal identity line.
We also use the `de` logical vector to highlight the significantly differentially expressed genes.

```{r expplot2}
exp <- marray$expression
plot(exp[, 1], exp[, 4],
     xlab = colnames(exp)[1],
     ylab = colnames(exp)[4])
title(main = "Comparing expression levels")
grid()
points(exp[de, 1], exp[de, 4], col = "red", pch = 19)
abline(0, 1)
```

We can identify individual points on the plot

```{r, eval = FALSE}
identify(exp[, 1], exp[, 4], labels = fmeta$genes)
```

The `pairs` function creates all possible pairs of scatter plots

```{r plotpairs}
pairs(exp)
```

We can also represent the data using density colouring instead of plotting 
individual points.

```{r smoothscatter}
smoothScatter(exp[, 1], exp[, 4])
```

A histogram of expression values for sample 1

```{r exphist}
hist(exp[, 1])
```

Below, we prepare a 3 by 2 matrix of figures with `par` and use a `for` loop 
to plot the 6 histograms of intensities.

```{r exphist6}
par(mfrow = c(3, 2))
for (i in 1:6) {
  hist(exp[, i],
       xlab = marray$samplemeta[i, "sample"],
       main = "Histrogram of intensities")
  rug(exp[, i])
}
```

By extracting the expression data, we have lost some information 
that was stored in the respective meta data elements. 
Let set the sample and gene names.

```{r expcolrownames}
colnames(exp) <- marray$samplemeta$sample
rownames(exp) <- marray$featuremeta$genes
```

Below, we plot a heatmap of all genes and those that were deemed significantly differentially expressed.

```{r expheatmaps}
heatmap(exp)
heatmap(exp[de, ])
```

## Counting number of DE genes

We first use the `dir` function to list all the files in our working directory 
and then filter only those with a pattern of interest.

```{r dirfmeta}
dir()
dir(pattern = "fmeta")
```

We could use `dir` to create a vector of file names of interest and 
iterate over those, as illustrated below.

```{r flsforloop}
for (fmetafile in dir(pattern = "fmeta")) {
  fmeta <- read.csv(fmetafile, row.names = 1)
  print(dim(fmeta))
}
```

Alternatively, we create these file names manually to illustrate the `paste0` function.
We then apply the same strategy as above to read the feature meta data into `R`, 
extract the differentially expressed genes and, if there are any, 
print how many have been counted, import the expression data and save
a heatmap into a pdf file.

```{r paste0forloop}
for (i in 1:3) {
  fmetafile <- paste0("fmeta", i, ".csv")
  cat(fmetafile, ":\n")
  fmeta <- read.csv(fmetafile, row.names = 1)
  de <- which(fmeta$bh < 0.05)
  if (length(de) > 0) {
    expfile <- paste0("MAdata", i, ".csv")
    exp <- read.csv(expfile, row.names = 1)
    exp <- as.matrix(exp)
    pdffile <- paste0("heatmap", i, ".pdf")
	cat(length(de), "DE genes.\n")
	cat("Saved", pdffile, ".\n")
    pdf(pdffile)
    heatmap(exp[de, ])
    dev.off()
  } else {
    cat("No DE found.\n")
  }
  cat("\n")
}
```
