# My first simple Rmd file

Don't forget to install knitr package.

```{r, eval=FALSE}
install.packages(c("knitr"))
```

# Introduction

This is an .Rmd document.  Maybe you have heard of Sweave?
Well, knitr is like Sweave, only better!

* can work with markdown files (simpler)
* more flexible re: graphics
* caching
* "modern"

Many of the notes in Rmd are supplied on the github pages for this
bootcamp.  We convert them to .md which github renders.

## A simple example

Let's get R to do some simple maths.

```{r}
x <- 1:3
y <- 4:6
outer(x, y)
```
As well as having chunks in separate paragaphs, you can have some
inline computation.  For example, the mean of x is `r mean(x)` and
the sum of x and y is `r x+y`.  Chunks can be named and then
referred to later.

## Plotting is easy too

```{r plot-chunk}
x <- seq(from=0, to=2*pi, length=1000)
y <- cos(3*x)
plot(x, y, type='l', col='blue')
```

## Tables are not so straightforward (in markdown)

See the
[xtable](http://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf)
package gallery for more advanced examples.

```{r,results='asis'}
kable(head(iris[,1:3]), format='html')
```


# Okay, what do you need to know?


1. What is markdown syntax?  [Markdown](http://daringfireball.net/projects/markdown)

2. What can knitr do? [Knitr home page](http://yihui.name/knitr/)


# Exercise:

Take a look at the example for estimating PI using a well-known method
of throwing darts at a dartboard.

[Estimating PI](http://www.damtp.cam.ac.uk/user/eglen/rguide/estimate.pdf)

Can you generate something similar with an .Rmd document?  Work on the
basics of estimating pi and attempt the last figure only if you have
time.

# What are vignettes?

Vignettes are self-contained documents that typically describe in
detail how a package works.  See a list of them with `vignette()`.
*knitr* has many vignettes, e.g. `vignette('knitr-refcard')`.

# Installing vignettes into a package.

knitr vignettes take a bit more work, but not too much.  Put them in
`PKG/inst/doc` or `PKG/vignettes/` folder and follow instructions.
e.g.  for markdown, see the file

```{r}
system.file('doc', 'knitr-markdown.Rmd', package='knitr')
```


# How to compile this document

This should work cross platform; first, the document is knitted to
make a markdown file; then the "markdown" package converts this to
html which can be rendered.

```{r, eval=FALSE}
require(knitr)
knit2html("simple.Rmd")
if (interactive()) browseURL("simple.html")
```

In R studio, it is even simpler; just hit the special "Knit HTML" button!

If you prefer the pdf format you need to have a latex compiler, so the
following may not work for you (especially on Windows):

```{r, eval=FALSE}
require(knitr)
knit2html('simple.Rmd'); browseURL('simple.html')
knit2pdf("simple.Rmd")
```

## About this document

```{r}
sessionInfo()
```
