## A vector of `integer`s

We are going to use the `sample` function to create such a vector. 
As its name implied, the function samples among a set it is provided as imput. 
It is as such important to provide a set in integers. As explained in 
`?sample`, if given a single numeric input `n`, `sample` will sample `n` values 
from `1:n` without replacement (`replacement = FALSE`), resulting in a permutation of `1:n`.

Using `n` equal `20` and making sure that `1:20` produces a sequence of integers

```{r}
class(1:20)
```

we can now produce our vector of integers `i`:

```{r}
i <- sample(20)
i
typeof(i)
```

## A vector of `double`s

The `runing` sample a certain number of values from a uniform distribution ranging (by default)
from 0 to 1. 

```{r}
d <- runif(20)
d
typeof(d)
```

We could also have used `rnorm` to sample values from a normal distribution. 

```{r}
typeof(rnorm(20))
```
## A vector of `character`s

Instead of typing the sequence of characters ourselves, we will use the 
built-in variable `letters`, which stores the 26 letters from the alphabet. 
We will select the letters at positions 1 to 20 in the order defined by `i`.

```{r}
head(letters)
s <- letters[i]
s
typeof(s)
length(s)
```

Other built-in variables of interest are `LETTERS` (as `letters`, but capital letters), 
`month.name` (month names), pi, ... see `?letters`.

## A vector of `logical`s

Although `TRUE` is a perfectly valid vector of logicals (of length 1), we are going to use 
a logical operator on the vector `i` to create a vector of logicals. Each element of `i` 
will be used to the comparison in turn, effectively resulting in a vector of logicals of the 
same length than `i`.

```{r}
l <- i >= 10
l
typeof(l)
length(i)
length(l)
```


