# How to store microarray data?

## The expression data

The most natural way to store the expresion data is a `matrix`,
where genes are represented along the rows and samples along the columns.
All values are of the same type (numerical values). 
Below we generate a matrix of random (normally distributed data),
representing 50 genes measured for 4 samples.

```{r expdata}
expdata <- matrix(rnorm(200), nrow = 50, ncol = 4)
head(expdata)
```

Note that this also support missing data encoded as `NA`.

```{r expdatana}
expdatana <- expdata
expdatana[2, 2] <- NA
head(expdatana)
class(expdatana)
mode(expdatana)
```

We also want to set gene (row) and sample (column) names.

```{r expnames}
dimnames(expdata) <-
  list(features = paste0("gene", 1:nrow(expdata)),
       samples = paste0("sample", 1:ncol(expdata)))
```

## The meta data

We provide addition information on the genes and samples using two data frames. 
We have to make sure that the respective dimensions of the expression data 
and gene/sample meta data match.

```{r smdata}
smdata <- data.frame(feature = colnames(expdata),
                     group = c("ctrl", "ctrl",
                       "cond1", "cond1"),
                     replicate = rep(1:2, each = 2))
smdata
class(smdata)
nrow(smdata)
ncol(expdata)
nrow(smdata) == ncol(expdata)
```

```{r fmdata}
fmdata <- data.frame(feature = rownames(expdata),                     
                     description = paste("Important gene",
                       rownames(expdata)))
fmdata
nrow(fmdata)
nrow(expdata)
nrow(fmdata) == nrow(expdata)
```

## A complete representation

We now combine the respective objects into a list to keep the 
expression data and its description togeter. 
We then print the structure of the list as a summary to ensure that 
we have the expected representation.


```{r marray}
marray <- list(
  expression = expdata,
  featuremeta = fmdata,
  samplemeta = smdata)
str(marray)
```
