# Another microarray analysis

## Combine expression data

This this, we use `dir` and a file pattern to create a vector 
of file names.

```{r}
fls <- dir(pattern = "Exercise-05-DE")
length(fls)
fls
```

We then use the `lapply` function that iterates over vector of file names 
(passed as first argument) and applies the `read.delim` function 
(the second argument) on each element of `fls`. The third argument of 
`lapply` is passed to `read.delim` to set the resulting `data.frame` row names
using the first column. As suggested by the letter `l`, `lapply`, returns 
a list of length equal to the length of its first argument. 
Because `read.delim` returns a `data.fame`, `l` will be a list of `data.frames`.

```{r}
l <- lapply(fls, read.delim, row.names = 1)
length(l)
class(l)
```

To obtain a vector of unique gene names, we iterate through the list 
of `data.frame`, extract their respective row names to create a list 
of characters (the gene names). We then `unlist`/flatten this list, 
producing a vector containing all the elements of the list and 
finall keep `unique` instances of each gene name (removing all duplicated gene names).


```{r}
allgenes <- lapply(l, rownames)
allgenes <- unique(unlist(allgenes))
length(allgenes)
head(allgenes)
```

We now know how many unique genes have been observed in total 
(`length(allgenes)`) and the number of experiments (`length(l)`) 
and initialise a matrix of adequate dimensions with 0s. We also 
use the gene names to set to matrix row names.

```{r}
eset <- matrix(0, nrow = length(allgenes), ncol = length(l))
rownames(eset) <- allgenes
```

We populate the matrix by iterating over the number of replicates. 
On each iteration, we `match` the position of the genes in our big
matrix and the individual replicate data stored in the `l` matrix. 
`idx` gives us the row indices of the exeriments in the big `eset` 
matrix. We then use these, and the loop index to set the expression 
data.

```{r}
for (i in 1:length(l)) {
  ## find the row indices
  idx <- match(rownames(l[[i]]), rownames(eset))
  eset[idx, i] <- l[[i]][, 1]
}
```

Finally, we plot a heatmap of `eset`.

```{r}
heatmap(eset)
```

##  Extracting and visualising genes of interest  

We have a table of validated genes that need to be parsed. 
We first read the data, assuring that the content of the 
resulting `data.frame` is interpreted as characters instead 
as factors.

```{r}
tab <- read.delim("Exercise-05-table.tsv", stringsAsFactors = FALSE)
tab
```

We then  split the comma separated genes names to produce a list 
(of length 2 because there were two rows in `tab`), which is then 
flattened with `unlist`.

```{r}
genes <- unlist( strsplit(tab$genes, ",") )
genes
```
Before subsetting our `eset` matrix, we check that all the extracted 
gene names are present.

```{r}
all(genes %in% rownames(eset))
```
```{r}
heatmap(eset[genes, ])
```
