# A short Bioconductor data analysis

We start by loading the packages that will provide the infrastrucutre 
to the Affymetric microarrys that we want to import and explore.

```{r lib, message = FALSE}
library("Biobase")
library("affy")
library("AnnotationDbi")
```

The Affymetrix raw data files are listed and saved in a vector of file names


```{r fls}
dir()
flnms <- dir(pattern = "*CEL")
flnms
```

All raw data files and then read with the specialised `ReadAffy` function
to produce a specific object of class `AffyBatch`.

```{r readraw}
rawdata <- ReadAffy(filenames = flnms)
class(rawdata)
```

Below, we show how to plot the array using the `image` function.

```{r plotraw}
par(mfrow = c(2, 2))
image(rawdata[, 1:4])
```

Let perform some quality control and assess whether RNA degradation is a concern here. 
The `AffyRNAdeg` function calculates a degradation pattern for each samples, 
that can then be plotted with `plotAffyRNAdeg`.

```{r rnadeg}
deg <- AffyRNAdeg(rawdata)
plotAffyRNAdeg(deg)
```

The `AffyBatch` object stores raw data. Before proceeding with data analysis, 
we normalise the data with the popular robust multi-array average methods, 
implemented in the `rma` function. This produces another type of specialised 
object, an `ExpressionSet`. We can extract the assay data with the `exprs` 
accessor.

```{r norm}
eset <- rma(rawdata)
class(eset)
eset
head(exprs(eset))
```

We can now produce a boxplot of normalised intensities

```{r boxplot}
boxplot(exprs(eset))
```

Let's extract all control probe sets that are labelled `AFFX`, 
extract their measured expression intensities and generate a 
heatmap.

```{r ctrlhmap}
ctrl <- grep("AFFX", featureNames(eset))
heatmap(exprs(eset[ctrl, ]))
```

The code below illustrates a principal component analysis. 
The samples are then plotted on a 2 dimensional PCA plot 

```{r pca}
pca <- prcomp(t(exprs(eset)))
plot(pca$x[, 1:2], pch = 19, col = "#AB000030", cex = 3)
text(pca$x[, 1:2], labels = 1:5)
grid()
```

We can save the `ExpressionSet` for later re-use with the 
`save` function.

```{r save, eval=FALSE}
save(eset, file = "eset.rda")
```

A complete set of quality metrics and plots are available through the 
`arrayQualityMetrics` function from the `arrayQualityMetrics` package. 
This function creates a complete html report in the `aqmReport` directory. 


```{r aqm, eval = FALSE}
library("arrayQualityMetrics")
arrayQualityMetrics(eset, outdir = "aqmReport")
```

See the vignette for more details.

```{r vig, eval = FALSE}
vignette(package = "arrayQualityMetrics")
vignette("arrayQualityMetrics", package = "arrayQualityMetrics")
```
