# The ABRF data

Bennett et al. *The 2012/2013 ABRF Proteomic Research Group Study:
Assessing Longitudinal Intralaboratory Variability in Routine Peptide
Liquid Chromatography Tandem Mass Spectrometry*
[Molecular & Cellular Proteomics, 14, 3299-3309](http://www.mcponline.org/content/14/12/3299.short).

> Questions concerning longitudinal data quality and reproducibility
> of proteomic laboratories spurred the Protein Research Group of the
> Association of Biomolecular Resource Facilities (ABRF-PRG) to design
> a study to systematically assess the reproducibility of proteomic
> laboratories over an extended period of time. Developed as an open
> study, initially 64 participants were recruited from the broader
> mass spectrometry community to analyze provided aliquots of a six
> bovine protein tryptic digest mixture every month for a period of
> nine months. Data were uploaded to a central repository, and the
> operators answered an accompanying survey. Ultimately, 45
> laboratories submitted a minimum of eight LC-MSMS raw data files
> collected in data-dependent acquisition (DDA) mode. No standard
> operating procedures were enforced; rather the participants were
> encouraged to analyze the samples according to usual practices in
> the laboratory. Unlike previous studies, this investigation was not
> designed to compare laboratories or instrument configuration, but
> rather to assess the temporal intralaboratory reproducibility. The
> outcome of the study was reassuring with 80% of the participating
> laboratories performing analyses at a medium to high level of
> reproducibility and quality over the 9-month period. For the groups
> that had one or more outlying experiments, the major contributing
> factor that correlated to the survey data was the performance of
> preventative maintenance prior to the LC-MSMS analyses. Thus, the
> Protein Research Group of the Association of Biomolecular Resource
> Facilities recommends that laboratories closely scrutinize the
> quality control data following such events. Additionally, improved
> quality control recording is imperative. This longitudinal study
> provides evidence that mass spectrometry-based proteomics is
> reproducible. When quality control measures are strictly adhered to,
> such reproducibility is comparable among many disparate groups. Data
> from the study are available via ProteomeXchange under the accession
> code PXD002114.

The 46 metrics generated by the 'IDFree' mode of QuaMeter.

```{r, echo = FALSE}
knitr::kable(read.csv("../data/20121213-ABRF-PRF-IDFree-metrics-feature-vars.csv"))
```

## Exercice

```{r, echo = FALSE}
echo. <- FALSE
```

* Locate the data file is `20121213-ABRF-PRF-IDFree-metrics.xlsx` on your disk.

```{r, echo = echo.}
idfile <- "../data/20121213-ABRF-PRF-IDFree-metrics.xlsx"
```

* It is an Excel file. We have seen how to read a text-based
  spreadsheet using `read.csv`, `read.delim`, `read.table`, ...  We
  can also use the `read_excel` function from the `readxl` package to
  read it in. Load the package in your R session.

```{r, echo = echo.}
library("readxl")
```
* In addition to the file name, we also need to provide which sheet to
  read the data from. Check the `read_xl` documentation page and
  identify how to do this.

```{r, eval = FALSE, echo = echo.}
?read_xl
```
* Read the data into R

```{r, echo = echo.}
idfree <- read_excel(idfile, sheet = "ID-free metrics")
```

* What are the dimensions of the data. Check the names of the
  columns. What class is the variable you created with the
  `read_excel` function.

```{r, eval = echo., echo = echo.}
dim(idfree)
colnames(idfree)
class(idfree)
```
