# My first simple Rmd file

Don't forget to install knitr package.

```{r, eval=FALSE}
install.packages(c("knitr"))
```

# Introduction

This is an .Rmd document.  Maybe you have heard of Sweave?
Well, knitr is like Sweave, only better!

* can work with markdown files (simpler)
* more flexible re: graphics
* caching
* "modern"

# A simple example

Let's get R to do some simple maths.

```{r}
x <- 1:3
y <- 4:6
outer(x, y)
```

# Plotting is easy too

```{r}
x <- seq(from=0, to=2*pi, length=1000)
y <- cos(3*x)
plot(x, y, type='l', col='blue')
```


# Okay, what do you need to know?


1. What is markdown syntax?  [Markdown](http://daringfireball.net/projects/markdown)

2. What can knitr do? [Knitr home page](http://yihui.name/knitr/)


# Exercise:

Take a look at the example for estimating PI using a well-known method
of throwing darts at a dartboard.

[Estimating PI](http://www.damtp.cam.ac.uk/user/eglen/rguide/estimate.pdf)

Can you generate something similar
with an .Rmd document?  Work on the basics of estimating pi and
attempt the last figure only if you have time.


# How to compile this document

This should work cross platform; first, the document is knitted to
make a markdown file; then the "markdown" package converts this to
html which can be rendered.

```{r, eval=FALSE}
require(knitr)
knit2html("simple.Rmd")
if (interactive()) browseURL("simple.html")
```

In R studio, it is even simpler; just hit the special "Knit HTML" button!

If you prefer the pdf format you need to have a latex compiler, so the
following may not work for you (especially on Windows):

```{r, eval=FALSE}
require(knitr)
knit2pdf("simple.Rmd")
```

## About this document

```{r}
sessionInfo()
```
