# R and Bioconductor for proteomics

## Bioconductor

[Bioconductor](http://www.bioconductor.org/) provides tools for the
analysis and comprehension of high-throughput biology (genomics,
proteomics, metabolomics, flow cytometry, ...)  data. Bioconductor
uses the R statistical programming language, and is open source and
open development. It has two releases each year, 1211 software
packages, and an active user community.

In addition to *software* package, it also has *experiment* and
*annotation* packages.

## Installing Bioconductor packages

First time

```{r, eval = FALSE}
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite("RforProteomics")
```

Then

```{r, eval = FALSE}
library(BiocInstaller)
biocLite("RforProteomics")
```

To install all Bioconductor software that will be used throughout this
workshop:

```{r, eval = FALSE}
biocLite(c("RforProteomics", "pRolocGUI"), dependencies = TRUE)
```

See the detailed session information in the [wrap up](../wrapup.md)
section for packages and versions used to generate this material.

## Proteomics and mass spectrometry

The easiest way to browser and search Bioconductor packages is to
navigate [task
views](http://www.bioconductor.org/packages/release/BiocViews.html#___Software).

Particularly relevant for this course are the [mass
spectrometry](http://www.bioconductor.org/packages/release/BiocViews.html#___MassSpectrometry),
[mass spectrometry
data](http://www.bioconductor.org/packages/release/BiocViews.html#___MassSpectrometryData)
and
[proteomics](http://www.bioconductor.org/packages/release/BiocViews.html#___Proteomics)
task views.

## The `RforProteomics` package

The
[`RforProteomics`](http://www.bioconductor.org/packages/release/data/experiment/html/RforProteomics.html)
package contains code to illustrate the *Using R and Bioconductor for
proteomics data analysis* and *Visualisation of proteomics data using
R and Bioconductor* papers. Two vignettes describe the code and data
needed to reproduce all the examples and figures.


```{r, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library("mzR"))
suppressPackageStartupMessages(library("MSnbase"))
```

```{r}
library("RforProteomics")
```

Package vignettes are overview package documentations. They generally
provide a complete use case demonstrating the package's
functionality. All Bioconductor software packages have vignettes in
addition to all function manuals.

## Proteomics/MS data structures

```{r, echo = FALSE}
knitr::kable(RforProteomics:::msDataTab())
```

## References

Gatto L. and Christoforou A. *Using R and Bioconductor for proteomics
data analysis*, Biochim Biophys Acta - Proteins and Proteomics,
2013. [PMID:23692960](https://www.ncbi.nlm.nih.gov/pubmed/23692960)
([preprint](https://arxiv.org/abs/1305.6559))

Gatto L, Breckels LM, Naake T, Gibb S. *Visualisation of proteomics
data using R and Bioconductor*. Proteomics. 2015 Feb 18. doi:
10.1002/pmic.201400392. [PMID:25690415](http://www.ncbi.nlm.nih.gov/pubmed/25690415).



