# Indentification data: `mzR`, `mzID`, `MSnID`

```{r env, echo=FALSE}
library(BiocStyle)
suppressPackageStartupMessages(library("mzR"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("MSnID"))
```
## Handling identification data

There are two packages that can be used to parse `mzIdentML` files,
namely `mzR` (that we have already used for raw data) and `mzID`. The
major difference is that the former leverages C++ code from
`proteowizard` and is hence faster than the latter (which uses the
`XML` R package). They both work in similar ways.

```{r, echo = FALSE}
x <- RforProteomics:::msDataTab()
sel <- x[, 1] %in% c("Identification")
knitr::kable(x[sel, ])
```

We are going to use the following identification file in this practical:

```{r}
library("msdata")
idf <- ident(full.names = TRUE)
basename(idf)
```

### mzID

The main functions are `mzID` to read the data into a dedicated data
class and `flatten` to transform it into a `data.frame`. 

```{r id}
library("mzID")
id <- mzID(idf)
id
```

Various data can be extracted from the `mzID` object, using one the
accessor functions such as `database`, `sofware`, `scans`, `peptides`,
... The object can also be converted into a `data.frame` using the
`flatten` function.

```{r fid}
head(flatten(id))
```

#### Exercise

Open the `r basename(idf)` file as shown above. What software and how many
proteins were in the database used to perform the search?

### `mzR`

The `mzR` interface provides a similar interface. It is however much
faster as it does not read all the data into memory and only extracts
relevant data on demand. It has also accessor functions such as
`softwareInfo`, `mzidInfo`, ... (use `showMethods(classes = "mzRident", where = "package:mzR")`)
to see all available methods.

```{r id2}
library("mzR")
id2 <- openIDfile(idf)
id2
softwareInfo(id2)
```

The identification data can be accessed as a `data.frame` with the
`psms` accessor.

```{r fid2}
head(psms(id2))
```

#### Example

Is there a relation between the length of a protein and the number of
identified peptides, conditioned by the (average) e-value of the
identifications?

```{r, ex_id, echo=TRUE, eval=TRUE}
fid <- flatten(id)
x <- by(fid, fid$accession,
        function(x)
            c(unique(x$length),
              length(unique(x$pepseq)),
              mean(x$'ms-gf:specevalue')))
x <- data.frame(do.call(rbind, x))
colnames(x) <- c("plength", "npep", "eval")
x$bins <- cut(x$eval, summary(x$eval))
library("lattice")
xyplot(log10(plength) ~ npep | bins, data = x)
```

## Adding identification data to raw data

Here are two matching raw and identiciation data files:

```{r}
library("MSnbase")
## find path to a mzXML file
rwf <- dir(system.file(package = "MSnbase", dir = "extdata"),
           full.name = TRUE, pattern = "mzXML$")
## find path to a mzIdentML file
idf <- dir(system.file(package = "MSnbase", dir = "extdata"),
           full.name = TRUE, pattern = "dummyiTRAQ.mzid")
```

We first create the raw data object:

```{r}
msexp <- readMSData(rwf, verbose = FALSE)
head(fData(msexp))
```

The simply add identification data. The matching of spectra from the
raw data and the PSMs from the identification data is done internally.

```{r}
msexp <- addIdentificationData(msexp, idf)
head(fData(msexp))
```

## Visualising identification data

For this part, let's use a ready made `MSnExp` object that is
distributed with the `MSnbase` package. Simply use the `data()`
function with the name of the desired data.

```{r}
library("MSnbase")
data(itraqdata)
```

### Annotated spectra and spectra comparison

```{r id1, message=FALSE, fig.width=15, message=FALSE}
par(mfrow = c(1, 2))
itraqdata2 <- pickPeaks(itraqdata, verbose = FALSE)
s <- "SIGFEGDSIGR"
plot(itraqdata2[[14]], s, main = s)
plot(itraqdata2[[25]], itraqdata2[[28]], sequences = rep("IMIDLDGTENK", 2))
```

The annotation of spectra is obtained by simulating fragmentation of a
peptide and matching observed peaks to fragments:

```{r fag}
calculateFragments("SIGFEGDSIGR")
```

Visualising a pair of spectra means that we can access them, and that,
in addition to plotting, we can manipulate them and perform
computations. The two spectra corresponding to the `IMIDLDGTENK`
peptide, for example have 

```{r}
compareSpectra(itraqdata2[[25]], itraqdata2[[28]], fun = "common")
```

common peaks, a correlation of

```{r}
compareSpectra(itraqdata2[[25]], itraqdata2[[28]], fun = "cor")
```

and a dot product of 

```{r}
compareSpectra(itraqdata2[[25]], itraqdata2[[28]], fun = "dotproduct")
```

See `?compareSpectra` for details.

There are 2 Bioconductor packages for peptide-spectrum matching
directly in R, namely `r Biocpkg("MSGFplus")` and `r Biocpkg("rTANDEM")`, 
replying on `MSGF+` and `X!TANDEM` respectively.
See also the `r Biocpkg("MSGFgui")` package for visualisation of
identification data.


## Exploration and Assessment of Confidence of LC-MSn Proteomics Identifications using `MSnID`

The `MSnID` package extracts MS/MS ID data from mzIdentML (leveraging
the `mzID` package) or text files. After collating the search results
from multiple datasets it assesses their identification quality and
optimises filtering criteria to achieve the maximum number of
identifications while not exceeding a specified false discovery
rate. It also contains a number of utilities to explore the MS/MS
results and assess missed and irregular enzymatic cleavages, mass
measurement accuracy, etc.

### Step-by-step workflow

Let's reproduce parts of the analysis described the `MSnID`
vignette. You can expole more with 

```{r, eval = FALSE}
vignette("msnid_vignette", package = "MSnID")
```

The `r Biocpkg("MSnID")` package can be used for post-search filtering
of MS/MS identifications. One starts with the construction of an
`MSnID` object that is populated with identification results that can
be imported from a `data.frame` or from `mzIdenML` files. Here, we
will use the example identification data provided with the package.

```{r idf}
mzids <- system.file("extdata", "c_elegans.mzid.gz", package="MSnID")
basename(mzids)
```

We start by loading the package, initialising the `MSnID` object, and
add the identification result from our `mzid` file (there could of
course be more that one).

```{r msnid1}
library("MSnID")
msnid <- MSnID(".")
msnid <- read_mzIDs(msnid, mzids)
show(msnid)
```

Printing the `MSnID` object returns some basic information such as

* Working directory.
* Number of spectrum files used to generate data.
* Number of peptide-to-spectrum matches and corresponding FDR.
* Number of unique peptide sequences and corresponding FDR.
* Number of unique proteins or amino acid sequence accessions and corresponding FDR.


The package then enables to define, optimise and apply filtering based
for example on missed cleavages, identification scores, precursor mass
errors, etc. and assess PSM, peptide and protein FDR levels. To
properly function, it expects to have access to the following data

```{r msnidcols, echo=FALSE}
sort(MSnID:::.mustBeColumns)
```

which are indeed present in our data:

```{r msnidnames}
names(msnid)
```

Here, we summarise a few steps and redirect the reader to the
package's vignette for more details:


### Analysis of peptide sequences

Cleaning irregular cleavages at the termini of the peptides and
missing cleavage site within the peptide sequences. The following two
function call create the new `numMisCleavages` and `numIrregCleavages`
columns in the `MSnID` object

```{r msnidtermini}
msnid <- assess_termini(msnid, validCleavagePattern="[KR]\\.[^P]")
msnid <- assess_missed_cleavages(msnid, missedCleavagePattern="[KR](?=[^P$])")
```

## Trimming the data

Now, we can use the `apply_filter` function to effectively apply
filters. The strings passed to the function represent expressions that
will be evaludated, this keeping only PSMs that have 0 irregular
cleavages and 2 or less missed cleavages.

```{r msnidtrim}
msnid <- apply_filter(msnid, "numIrregCleavages == 0")
msnid <- apply_filter(msnid, "numMissCleavages <= 2")
show(msnid)
```

## Parent ion mass errors

Using `"calculatedMassToCharge"` and `"experimentalMassToCharge"`, the
`mass_measurement_error` function calculates the parent ion mass
measurement error in parts per million.

```{r msnidppm1}
summary(mass_measurement_error(msnid))
```

We then filter any matches that do not fit the +/- 20 ppm tolerance

```{r msnidppm2}
msnid <- apply_filter(msnid, "abs(mass_measurement_error(msnid)) < 20")
summary(mass_measurement_error(msnid))
```

## Filtering criteria

Filtering of the identification data will rely on 

* -log10 transformed MS-GF+ Spectrum E-value, reflecting the goodness
  of match experimental and theoretical fragmentation patterns
  
```{r filt1}
msnid$msmsScore <- -log10(msnid$`MS-GF:SpecEValue`)
```  

* the absolute mass measurement error (in ppm units) of the parent ion

```{r filt2}
msnid$absParentMassErrorPPM <- abs(mass_measurement_error(msnid))
```

## Setting filters

MS2 fiters are handled by a special `MSnIDFilter` class objects, where
individual filters are set by name (that is present in `names(msnid)`)
and comparison operator (>, <, = , ...)  defining if we should retain
hits with higher or lower given the threshold and finally the
threshold value itself.

```{r filt3}
filtObj <- MSnIDFilter(msnid)
filtObj$absParentMassErrorPPM <- list(comparison="<", threshold=10.0)
filtObj$msmsScore <- list(comparison=">", threshold=10.0)
show(filtObj)
```

We can then evaluate the filter on the identification data object,
which return the false discovery rate and number of retained
identifications for the filtering criteria at hand.

```{r filt4}
evaluate_filter(msnid, filtObj)
```

## Filter optimisation

Rather than setting filtering values by hand, as shown above, these
can be set automativally to meet a specific false discovery rate.

```{r optim1}
filtObj.grid <- optimize_filter(filtObj, msnid, fdr.max=0.01,
                                method="Grid", level="peptide",
                                n.iter=500)
show(filtObj.grid)
```

```{r optim2}
evaluate_filter(msnid, filtObj.grid)
```

Filters can eventually be applied (rather than just evaluated) using
the `apply_filter` function.

```{r optim3}
msnid <- apply_filter(msnid, filtObj.grid)
show(msnid)
```

And finally, identifications that matched decoy and contaminant
protein sequences are removed 

```{r optim4}
msnid <- apply_filter(msnid, "isDecoy == FALSE") 
msnid <- apply_filter(msnid, "!grepl('Contaminant',accession)")
show(msnid)
```

## Export `MSnID` data

The resulting filtered identification data can be exported to a
`data.frame` or to a dedicated `MSnSet` data structure for
quantitative MS data, described below, and further processed and
analyses using appropriate statistical tests.

```{r}
head(psms(msnid))
as(msnid, "MSnSet")
```
