# Introduction

## Pre-requisites


## `R` is

### a dynamic language

```{r}
a <- 1L
mode(a)
typeof(a)

b <- 1
mode(b)
typeof(b)

x <- "1"
mode(x)
typeof(x)

paste(a, x)
```
### a functional programming language 

Functions are *first class citizens*. Use function (outputs) as inputs
to other variables, to create other variables, ... we will make use of
this throughout the workshop. More details
[here](https://github.com/lgatto/R-functional-programming#readme).

### lazy

Calling a function with argument *wait for 3 seconds* takes no time to return
```{r}
f <- function(x) { 10 }
system.time(f(Sys.sleep(3)))
```

unless we *force* evaluation

```{r}
f <- function(x) { force(x); 10 }
system.time(f(Sys.sleep(3)))
```

(Example originally from Hadley Wickham's devtools page)

### object-oriented

with multiple frameworks: S3, S4 and S4 reference classes

## Data structures

Introduce data structures and manipulation 

- vectors
- numerics, characters, logicals, factors
- matrix
- array
- list
- data.frame
- special values: `NULL`, `NA`, `NaN`

### Objects

As in OO programming:

```{r}
class(x <- rnorm(100))
y <- rnorm(100)
class(1L)
class("123")
class(sum)
class(lm(y ~ x))
```

