R programming
===

```{r, echo=FALSE}
opts_chunk$set(eval=FALSE)
```

# IO

## Reading text spreadsheets

Using `read.table` *et al.*, a text spreadsheet (in `csv`, `tsv` or
similar) can be read in and converted into a `data.frame`. As usual,
text columns as converted into factors unless `stringsAsFactor=FALSE`
or `as.is=TRUE`.

```{r}
f <- "2014_01_01"
lf <- file.path("../data/daily-text", f)
w <- read.table(lf,
                header = FALSE,
                comment.char = "#",
                sep = "\t")
dim(w)
head(w)
```

We are missing header information. Let's use `readLines` to parse and
extract the header data manually. 

```{r}
hd <- readLines("../data/daily-text/2014_01_01")
hd <- hd[grep("#", hd)]
hd <- sub("#", "", hd)
hd <- hd[7:8]
hd <- gsub(" ", "", hd)
hd <- strsplit(hd, "\t")
hd <- paste0(hd[[1]], " [", hd[[2]], "]")
hd <- sub(" \\[\\]", "", hd)
names(w) <- hd
```

The format of the first columns `Time` is still unsatisfactory. Below,
we convert it into a time format/date.

```{r}
class(w$Time)
w$Time <- strptime(paste(f, w$Time), "%Y_%m_%d %H:%M")
w$Day <- as.Date(f, "%Y_%m_%d")
class(w$Time)
summary(w)
```

## Basic plotting

```{r}
par(mfrow = c(2, 2))
plot(w$Time, w[, "Temp [degC]"], type = "b", xlab = "Time", ylab = "Temp")
plot(w$Time, w[, "WindSp [knots]"], type = "b", xlab = "Time", ylab = "Wind speed")
plot(w$Time, w[, "Rain [mm]"], type = "b", xlab = "Time", ylab = "Rain")
plot(w$Time, w[, "Press [mBar]"], type = "b", xlab = "Time", ylab = "Pressure")
```

```{r}
boxplot(w[, "WindSp [knots]"] ~ factor(w$WindDr))
pairs(w[, c(2, 5, 6, 9)])
```

## More plotting

While using two axes can be very misleading when the scales are
different (as in the example below) and the differences are not
properly accounted for, let's illustrate such an example to learn how
to set different elements of a base plot. 

- Data re-scaling

```{r}
temp0 <- w[, "Temp [degC]"]
temp <- temp0 - min(temp0) ## min is 0
temp <- temp/max(temp) ## max is 1

press0 <- w[, "Press [mBar]"]
press <- press0 - min(press0)
press <- press/max(press)
```

- Plot with minimal decoration

```{r}
par(mar = c(5, 4, 2, 4))
plot(w$Time,  temp , type = "l",
     xlab = "Time", ylab = "Temp [deg C]",
     yaxt = "n", col = "steelblue")
lines(w$Time, press, col = "red")
```

- Axis, title and legends

```{r}
axis(2, at = seq(0, 1, length = 11),
     labels = seq(min(temp0),
         max(temp0),
         length = 11))
axis(4, at = seq(0, 1, length = 11),
     labels = seq(min(press0),
         max(press0),
         length = 11))
mtext("Pressure [mBar]", 4, line = 3)
title(f)
legend("top", c("Temperature", "Pressure"),
       col = c("steelblue", "red"), lty = 1,
       bty = "n")
```

### Exercise

Using a weather data frame as input, generate a plot of the
temperature throughout the day.


## Writing text spreadsheets

`write.table` and `writeLines` to write tables and lines to files.

```{r, eval=FALSE}
write.csv(w, file = "w.csv")
```

## Saving and loading R data

Once data has been loaded an properly formatted into `R`, the easiest
and fastest way to serialise it is with `save`. Once stored in binary
format, the saved variable can be loaded in the working environment
with `load`. This can be used for one, several or all object in the
workspace (see `save.image`).

To save and restore single object to a file, see also `saveRDS` and
`readRDS`.

### Exercise

Use one of the `save`/`load` or `saveRDS`/`readRDS` above to serialise
the weather `data.frame`, rename the variable in your workspace,
reload the saved object and compare it to the original file. Hint: use
`identical` or `all.equal`.

### Bonus

```{r}
x <- sqrt(2)
x^2 == 2
all.equal(x^2, 2)
```
See [Why doesn’t R think these numbers are equal?](http://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f).

## See also
- `scan`: Read data into a vector or list from the console or file.
- User input: `menu`

```{r}
choices <- c("RStudio", "Wordpad", "emacs", "vim", "Notepad++")
mychoice <- menu(choices, graphics = FALSE, title = "Best editor ever")
cat("Best editor ever is ", choices[mychoice], "\n")
```
- Database: `RMySQL`, `RMongoDB`, `ROracle`
- `rhdf5`, `ncdf`, `XML`, `RJSONIO`, `jsonlite`
- web: `RCurl`, `httr`

# Iteration and flow control 

## Iteration

Iteration works as in other procedural languages

- `for (var in seq) expr`
- `while (cond) expr`
- `repeat expr`

where

- `expr` is a valid R expression, contained in `{` `}` if composed by
  multiple expressions.
- `cond` is a `logical` of length, or an expression that is evaluated
  to a logical. (It is also possible to use a `numeric` of length 1,
  where 0 corresponds to `FALSE` and any other number to `TRUE`).
- `seq` is a `vector` or a `list`.
- `var` is a variable that takes the values of `seq` throughout the
  iterations and can be used in `expr`.

Examples:

```{r}
for (i in 1:3)
    print(i + 1)

k <- 3
while (k > 0) {
    print(k^2)
    k <- k - 1
}
```

A common use of `for` is `for (i in 1:n)`, where `n` is the length of
a `vector` or a `list`. In such situations, it is safer to use
`seq_len(n)`:

```{r}
x <- numeric()
n <- length(x)

for (i in 1:n)
    print(i)

for (i in seq_len(n))
    print(i)
```

## Conditions

`if (cond) expr1 else expr2`, with `cond` and `expr[1|2]` as defined above. The else clause is optional.

It is also possible to nest `else/if` conditions:

```
if (cond1) {
	do1
} else if (cond2) {
	do2
} else {
	do3
}
```

Example:

```{r}
k <- 3
repeat {
    print(k)
    if (k == 0) break
    k <- k - 1
}
```

## `ifelse`

The vectorised `ifelse` function takes three expressions as arguments:
`test`, `yes` and `no`. All values of the `test` expression that
evaluate to `TRUE` are replaced by the corresponding value of `yes`
(possibly repeated) or the corresponding value of `no` otherwise.

```{r}
x <- 1:5
ifelse(x < 3, 1, 2)
ifelse(x < 3, 10:15, 20:25)
ifelse(x < 3, 10:15, -1)
```
If `test` has dimensions, there will be retained in the output:

```{r}
m <- matrix(1:6, ncol = 2)
m
ifelse(m < 2, m, c(10, 11, 12))
```

## `switch`

`switch`'s first argument is an expression, followed by any number of
additional arguments.

- If the first expression is evaluated to a numeric, the corresponding
  following argument is evaluated and returned.

```{r}
switch(1+0, 1, 2, 3)
switch(1+1, 1, 2, 3)
```

- If the expression evaluates to a character, the matching name of the
  following arguments is evaluated and returned.

```{r}
switch(letters[1+0], a = 1, b = 2, c = 3)
switch(letters[1+2], a = 1, b = 2, c = 3)
```

See `?switch` for details about mixing named and unnamed supplementary
arguments and partial matching.

We will see another `switch` example below, in the section about
functions.

## `*apply` and friends

When one wants to apply a function over the elements/dimensions of an
object:

| function | input | output | 
|----------|-------|--------|
| sapply   | list/vector  | vector | 
| lapply   | list/vector  | list   |
| apply    | matrix/array | vector/matrix [*] |

[*] will actually depend on the subscripts the function will applied
over.



- Retrieve the length of each element of a `list`: `sapply` the function `length` on each element of the list.

```{r}
l <- list(1:4, letters, month.name)
sapply(l, length)
```
- Same as above, but returning a list.

```{r}
lapply(l, length)
```

- Lets generate 3 sequences from 1 to 3, 4, and 5: apply function
  `seq_len` along the vector `c(3, 4, 5)`, to call `seq_len(3)`,
  `seq_len(4)` and `seq_len(5)`. Below, as the output can not be
  returned as a simple vector, the output is a `list`, even when using
  `sapply`.


```{r}
lapply(c(3, 4, 5), seq_len)
sapply(c(3, 4, 5), seq_len)
```

- Let's calculate the median of the rows and columns of a matrix:
  `apply` the function `median` over the rows (margin 1) and columns
  (margin 2) of `m`. Below, we generate a matrix of dimensions 6 by 5
  for 30 values sampled from a normal distribution `N(0, 1)` (see
  `?rnorm` for details).

```{r}
m <- matrix(rnorm(30), nrow = 6)
apply(m, 1, median)
apply(m, 2, median)
```

If we wanted to do the same thing with the `mean`, we would rather
want to use `rowMeans` and `colMeans`, which are much faster. See
later for details.

## Other `apply` functions

- `vapply`: same as `sapply`, but with pre-specified type of the
  return value.
- `mapply`: a multivariate version of `sapply` that applies a function
  `FUN` using the set of arguments passed to `apply` as function
  arguments to `FUN`.

```{r}
mapply(rep, 1:4, 4:1)
```
- `tapply`: applies a function to each cell of a ragged array. Below,
  sums the values of `x` after sub-setting them based on grouping
  defined in `k`. (See also `by` below).

```{r}
x <- 1:12
k <- rep(letters[1:3], 4)
tapply(x, k, sum)
```

- `rapply`: recursive version of `lapply`.

## Similar functions

- `replicate`: repeats the evaluation of an expression. 
- `aggregate`: splits the data into subsets, computes summary
  statistics for each, and returns the result in a convenient form.
- `by`: similar that `tapply` for data frames.

Reference:
[R Grouping functions](http://stackoverflow.com/questions/3505701/r-grouping-functions-sapply-vs-lapply-vs-apply-vs-tapply-vs-by-vs-aggrega)
on Stack Overflow.

## The `plyr` package

The [`plyr`](http://plyr.had.co.nz/) package provides its own set of
apply like functions. The `plyr` functions follows a simple naming
convention: `XYply` where `X` and `Y` describe the input and output
structures respectively, and can replaced by `a` (for an `array`), `l`
(for a `list`) or `d` (for a `data.frame`).

Good reference: [The Split-Apply-Combine Strategy for Data Analysis](http://www.jstatsoft.org/v40/i01).

## `for` or `apply`

The `apply` family of functions are not faster than for loops as long
as initialisation is accounted for (see section on benchmaking for
details). Their main benefits are conciseness and straightforward
parallelisation (next section).

Reference: [R Help Desk article (May 2008)](http://cran.r-project.org/doc/Rnews/Rnews_2008-1.pdf)

## Parallel `apply`

The `parallel` package provides a direct parallel alternatives for
`apply` functions with `mclapply`, `mcmapply`, `parLapply`,
`parSapply`, `parApply`, ...

Reference:
- `parallel` vignette: `vignette("parallel")`
- CRAN Task View: [High-Performance and Parallel Computing with R](http://cran.r-project.org/web/views/HighPerformanceComputing.html)
- Book: [Parallel R](http://shop.oreilly.com/product/0636920021421.do)
- The [`foreach`](http://cran.r-project.org/web/packages/foreach/index.html) package

<!-- ## TODO -->
<!-- - `melt`, `reshape2` -->
<!-- - `with` -->

# Writing function

Writing functions is very easy and a recommended way for code
abstraction. To create a new function, one needs to define:

1. A **name** that will be used to call the function (but see
   anonymous functions later); in the code chunk below, we call our
   function `myfun`.
2. A set of input formal **arguments**, that are defined in the
   parenthesis of the function constructor. The `myfun` example has
   two arguments, called `x` and `y`.
3. A function **body** (its code), defined between `{` and `}` below.
4. A **return** statement, that represents the output of the
   function. If no explicit return statement is provided, the last
   statement of the function is return by default.

```{r}
myfun <- function(x, y) {
    a <- x^2
    b <- sqrt(y)
    res <- a/b
    return(res)
}

myfun(2, 1)
myfun(4, 2)
```

Note that functions only support single value returns, i.e. `return(x,
y)` is an error.  To return multiple values, one needs to return a
`list` with the respective return variables like `return(list(x, y))`.


Calling `myfun(1)` would fail because argument `y` is not assigned a
value. One can assign default argument values when defining the function.


```{r}
myfun2 <- function(x, y = 2) {
    a <- x^2
    b <- sqrt(y)
    res <- a/b
    return(res)
}

myfun2(4, 2)
myfun2(4)
```

Note that the functions above are vectorised (they work with vectors
of arbitrary lengths), as their body is composed entirely if
vectorised functions.

```{r}
myfun(c(4, 4), c(1, 2))
```

## A example with `switch`

```{r}
centre <- function(x, type) {
       switch(type,
              mean = mean(x),
              median = median(x),
              trimmed = mean(x, trim = .1))
     }
x <- rcauchy(10)
centre(x, "mean")
centre(x, "median")
centre(x, "trimmed")
```

## The `...` argument

When an arbitrary number of arguments is to be passed to a function
(see `?cat` or `?rm` for examples) or if some arguments need to be
passed down to an inner function, one can use the special `...`
arguments.

```{r}
plot1toN <- function(n, ...) plot(1, n, ...)
```

### Exercise

- Write a function that takes a date character of the for
  `"YYYY-MM-DD"` as input and produces a plot of temperature over
  time. Make sure that it remains possible to fully customise the
  plot.

## Anonymous functions

It is often handy to define functions on the fly, without binding them
to specific names (item 1. above missing). These are called anonymous
functions and are generally used as one-time arguments to `apply`
functions.

```{r}
m <- matrix(rnorm(12), ncol = 4)
apply(m, 1, function(x) sum(x^2))
```

### Exercise

- Select all the weather files from June 2013. All file names are
  available with the `weatherfiles()` function. You can use the `grep`
  function to select the relevant file names. Check that you obtain 30
  files.

<!-- ```{r} -->
<!-- f <- grep("2013_06", fls, value = TRUE) -->
<!-- ``` -->

- Load the 30 data frames into a convenient data structure. Check the
  number of data points that are available for each weather data set.

<!-- ```{r} -->
<!-- xx <- lapply(f, weatherdata) -->
<!-- ``` -->

- Calculate the average day temperatures for that month.

<!-- ```{r} -->
<!-- sapply(xx, function(x) mean(x[, "Temp [degC]"])) -->
<!-- ## or -->
<!-- dd <- do.call(rbind, xx) -->
<!-- tapply(dd[, 2], dd$Day, mean) -->
<!-- ``` -->

## Scoping

In addition to what we have seen above, a function has also its very
own environment, in which it arguments a stored and its body is
evaluated. The functions arguments are copies of the initial
variables, so that the original ones stay unchanged.

```{r}
x <- 1
f <- function(x) {
    x <- x + 1
    return(x)
}
f(x)
x ## unchanged
```

A functions however can access variables defined outside of their
environment, for instance variables in the global environment.

```{r}
g <- function() {
    x <- x + 1
    return(x)
}
x <- 1
g()
x ## unchanged
```

In this case, if `x` does not exists

```{r}
rm(x)
g()
```

# R development

## Documentation

`R` features a dedicated documentation format (`Rd` files) that is
used when developing packages. It is extremely useful to document
code, even when the development of a package is not (yet) on the
agenda.

Comments are of course the first way to include context and additional
information to code. Code comments should describe why some code is
written and requirement for its correct execution and not what the
code does (which should be self-explanatory when reading it). 

The
[`roxygen2`](http://cran.r-project.org/web/packages/roxygen2/index.html)
package allows in-line documentation, i.e. documentation of `R` code
directly next (actually on top) to its source. This makes maintenance
of the code and the documentation reasonably easy and will allow to
automatically generate the `Rd` files at a later stage. 

```{r}
##' A short description of the function. Computes the ratio between
##' the square of \code{x} and the square-root of \code{y}. 
##'
##' Optional details section.
##' @title A simple function
##' @param x A vector of numerics.
##' @param y A vector of numerics.
##' @return A vector of numeric of length max(length(x), length(y)).
##' @author Laurent Gatto <lg390@@cam.ac.uk>
##' @examples
##' myfun(8, 3)
##' myfun(c(8, 4), c(3, 1))
myfun <- function(x, y) {
    a <- x^2
    b <- sqrt(y)
    res <- a/b
    return(res)
}
```

## Timing and benchmarking

To sample the execution time of a function, it is convenient to use
use `system.time` in conjunction with `replicate` and compute a
summary of the timings.

```{r}
X <- rnorm(1e6)
f <- function(x, k = .8) mean(x, trim = k)
f(X)
system.time(f(X))
summary(replicate(10, system.time(f(X))["elapsed"]))
```

Alternatively, the
[`rbenchmark`](http://cran.r-project.org/web/packages/rbenchmark/index.html)
and
[`microbanchmark`](http://cran.r-project.org/web/packages/microbenchmark/)
provide more formal benchmarking infrastructure. Let compare three
functions that create a list of length `n` composed of `1`, `1:2`, ..., `1:n`.

1. `f1` uses a for `loop` and grows the list dynamically at each
   iteration.
2. `f2` initialises the list and uses a `for` loop.
3. `f3` uses `lapply`.

```{r}
n <- 1e3
f1 <- function(n) {
    l <- list()
    for (i in seq_len(n))
        l[[i]] <- seq(i)
    return(l)
}

f2 <- function(n) {
    l <- vector("list", length = n)
    for (i in seq_len(n))
        l[[i]] <- seq(i)
    return(l)
}

f3 <- function(n) 
    lapply(seq_len(n), seq)
```

Let's use the `rbenchmark` package to compare the respective timings:

```{r}
library("rbenchmark")
benchmark(f1(n), f2(n), f3(n),
          columns = c("test", "replications", "elapsed", "relative"),
          replications = 5)
```

We see that the `for` with initialisation and `lapply` implementations
have comparable timings. The first function, however, takes much more
time. This overhead is the result of repeated copies of the list at
each iteration: before creating `l` of length `i`, the list of length
`i-1` is copied and deleted upon creation of the longer copy.

**Exercise:** write a parallel version of `f3` using `mclapply` using 2
cores. Do you see a 2-fold increase in speed?

For more extensive code profiling, see `?Rprof`.

## Debugging

To debug a function `f`, register is with `debug(f)`. Next time it is
called, it will be executed in `browser` mode: expressions of the body
can be executed one by one and at each step, the variables and their
values can be inspected.

<!-- ## `message`, `warning`, `error` -->
