R programming
===

```{r, echo=FALSE}
opts_chunk$set(eval=FALSE)
```

# IO

## Reading text spreadsheets

Using `read.table` *et al.*, a text spreadsheet (in `csv`, `tsv` or
similar) can be read in and converted into a `data.frame`. As usual,
text columns as converted into factors unless `stringsAsFactor=FALSE`
or `as.is=TRUE`.

```{r}
f <- "2014_01_01"
lf <- file.path("../data/daily-text", f)
w <- read.table(lf,
                header = FALSE,
                comment.char = "#",
                sep = "\t")
dim(w)
head(w)
```

We are missing header information. Let's use `readLines` to parse and
extract the header data manually. 

```{r}
hd <- readLines("../data/daily-text/2014_01_01")
hd <- hd[grep("#", hd)]
hd <- sub("#", "", hd)
hd <- hd[7:8]
hd <- gsub(" ", "", hd)
hd <- strsplit(hd, "\t")
hd <- paste0(hd[[1]], " [", hd[[2]], "]")
hd <- sub(" \\[\\]", "", hd)
names(w) <- hd
```

The format of the first columns `Time` is still unsatisfactory. Below,
we convert it into a time format/date.

```{r}
class(w$Time)
w$Time <- strptime(paste(f, w$Time), "%Y_%m_%d %H:%M")
class(w$Time)
summary(w)
```

## Plotting

```{r}
par(mfrow = c(2, 2))
plot(w$Time, w[, "Temp [degC]"], type = "b", xlab = "Time", ylab = "Temp")
plot(w$Time, w[, "WindSp [knots]"], type = "b", xlab = "Time", ylab = "Wind speed")
plot(w$Time, w[, "Rain [mm]"], type = "b", xlab = "Time", ylab = "Rain")
plot(w$Time, w[, "Press [mBar]"], type = "b", xlab = "Time", ylab = "Pressure")
```

```{r}
boxplot(w[, "WindSp [knots]"] ~ factor(w$WindDr))
pairs(w[, c(2, 5, 6, 9)])
```

## Writing text spreadsheets

`write.table` and `writeLines` to write tables and lines to files.

## Saving and loading R data

Once data has been loaded an properly formatted into `R`, the easiest
and fastest way to serialise it is with `save`. Once stored in binary
format, the saved variable can be loaded in the working environment
with `load`. This can be used for one, several or all object in the
workspace (see `save.image`).

To save and restore single object to a file, see also `saveRDS` and
`readRDS`.

### Exercise

Use one of the `save`/`load` or `saveRDS`/`readRDS` above to serialise
the weather `data.frame`, rename the variable in your workspace,
reload the saved object and compare it to the original file. Hint: use
`identical` or `all.equal`.

### Bonus

```{r}
x <- sqrt(2)
x^2 == 2
all.equal(x^2, 2)
```
See [Why doesn’t R think these numbers are equal?](http://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f).

## See also
- `scan`: Read data into a vector or list from the console or file.
- User input: `menu`

```{r}
choices <- c("RStudio", "Wordpad", "emacs", "vim", "Notepad++")
mychoice <- menu(choices, graphics = FALSE, title = "Best editor ever")
cat("Best editor ever is ", choices[mychoice], "\n")
```
- Database: `RMySQL`, `RMongoDB`, `ROracle`
- `rhdf5`, `ncdf`, `XML`, `RJSONIO`, `jsonlite`
- web: `RCurl`, `httr`

# Iteration and flow control 

- `for` and `while`
- `if`, `else`
- `*apply`
- `plyr`, `reshape2`
- `parallel`
- `replicate`
- `with`

# Writing function
- `function`
- pass by value (vs by reference)
- `environment`
- scoping

# Documentation

- `roxygen2` package syntax for in-line documentation

# Misc

- string processing: `strsplit`, `sub`, `gsub`, `paste`, `cat`
- more regexp: `stringr`, `tm`
- `message`, `warning`, `error`
- timing, benchmarking
- debugging
